Picky with peakpicking: assessing chromatographic peak quality with simple

### 2023-05-27 manuscript/manuscript.Rmd
Updating abstract to implement Anitra's comments on it

Chromatographic peakpicking continues to represent a significant bottleneck in automated LC-MS workflows. Uncontrolled false discovery rates and the lack of manually-calibrated quality metrics require researchers to visually evaluate individual peaks, requiring large amounts of time and breaking replicability. Here, we manually labeled two oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics and constructed a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis.

@@

A simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 80-90% down to 1-5% and showed minimal overfitting when applied to novel datasets. We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis, concluding that the poor performance of peakpicking algorithms significantly reduces the power of both univariate and multivariate statistical analyses in detecting environmental differences. We finally show that depth is a major driver of variability in the metabolome in high-quality environmental metabolomics datasets and identify several interesting metabolite groups for future investigation.

>>>>

Chromatographic peakpicking continues to represent a significant bottleneck in automated LC-MS workflows. Uncontrolled false discovery rates and the lack of manually-calibrated quality metrics require researchers to visually evaluate individual peaks, requiring large amounts of time and breaking replicability. This problem is exacerbated in noisy environmental datasets and for novel separation methods such as hydrophilic interaction columns in metabolomics, creating a demand for a simple, intuitive, and robust metric of peak quality.

@@

Here, we manually labeled four HILIC oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics and constructed a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis. We found that a simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 80-90% down to 1-5% and showed minimal overfitting when applied to novel datasets. We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis, concluding that the poor performance of peakpicking algorithms significantly reduces the power of both univariate and multivariate statistical analyses in detecting environmental differences. We finally show that depth is a major driver of variability in the metabolome and identify several interesting metabolite groups for future investigation.
